---
title: "STI Analysis Assignment"
author: "Kamau Nthoki"
format: html
embed-resources: true
editor: visual
---

# 1. Cleaning the data

```{r}
##working directory
require(haven)
setwd("C:/Users/Lilly Kamau/Desktop/RStudio")
getwd()
list.files()
stidata<-read.csv("STIData.csv")
```

```{r}
##cleaning data
summary(stidata)
##checking for missing values
colSums(is.na(stidata))
sum(is.na(stidata$Height ))
stidata <- stidata[!is.na(stidata$Height ), ]
# Replacing missing heights with the median height
summary(stidata$Height )
stidata$Height [is.na(stidata$Height )] <- 160
```

```{r}
##checking for duplicates
dups=duplicated(stidata$IdNumber)
stidata$IdNumber[duplicated(stidata$IdNumber)]
View(stidata[stidata$IdNumber==51,])
stidata<-stidata[!(stidata$IdNumber==51 & stidata$A1Age==23),]
View(stidata[stidata$IdNumber==51,])
```

```{r}
##checking for inconsistencies
summary(stidata)
str(stidata)
table(stidata$Sex )
for (col in names(stidata)) {
  cat("Unique values in", col, ":\n")
  print(unique(stidata[[col]]))
  cat("\n-----------------------\n")
}
stidata$N11Usedcondom <- tolower(stidata$N11Usedcondom)
stidata$N12UseCondom  <- tolower(stidata$N12UseCondom)
stidata$N13TakenAlcohol <- tolower(stidata$N13TakenAlcohol)
stidata$N10givereceiveforsex <- tolower(stidata$N10givereceiveforsex)
stidata$Sex <- tolower(stidata$Sex)

stidata$Sex <- trimws(stidata$Sex)
table(stidata$Height)
```

```{r}
##creating categorical variables
#Agecategory 
dim(stidata)
stidata$AgeCat<-NA
stidata$AgeCat[stidata$A1Age >= 16 & stidata$A1Age <= 34] <- 1
stidata$AgeCat[stidata$A1Age >= 35 & stidata$A1Age <= 64] <- 0
stidata$AgeCat <- factor(stidata$AgeCat, levels = c(1, 0), labels = c("16–34 yrs", "35–64 yrs"))
table(stidata$AgeCat)
#creating BMI 
stidata$BMI <- stidata$Weight / ((stidata$Height / 100)^2)
stidata$BMI <- cut(stidata$BMI,
                       breaks = c(0, 18.5, 24.9, 29.9, 34.9, Inf),
                       labels = c("Underweight", "Normal", "Overweight", "Obese Class I", "Obese Class II+"),
                       right = TRUE)
summary(stidata$BMI)

#Creating occupation
stidata$Occupation<-NA
stidata$Occupation[stidata$A2Occupation =="1 unemployed"|
                     stidata$A2Occupation == "4 student"]=1
stidata$Occupation[stidata$A2Occupation =="2 informal"|
                     stidata$A2Occupation =="3 formal"]=0
stidata$Occupation<-factor(stidata$Occupation, levels = c(1,0),labels = c("Unemployed", "Employed"))
summary(stidata$Occupation)

#creating churuch
class(stidata)
table(stidata$A3Church)
stidata$Religion<-NA
stidata$Religion[stidata$A3Church=="7 roman catholic"|
                   stidata$A3Church=="2 apostolic"|
                   stidata$A3Church=="3 methodist"|
                   stidata$A3Church=="4 anglican"|
                   stidata$A3Church=="5 pentecostal"]=1
stidata$Religion[ stidata$A3Church=="6 atheist"|
                    stidata$A3Church=="8 other"]=0

stidata$Religion<-factor(stidata$Religion, levels = c(1,0),labels = c("Belivers", "Non-Belivers"))
table(stidata$Religion)

##creating marital status
table(stidata$A5MaritalStatus)
class(stidata$A5MaritalStatus)
stidata$Marital_status<-NA
stidata$Marital_status[stidata$A5MaritalStatus=="2 married"]=1
stidata$Marital_status[stidata$Marital_status=="1 single"|
                                 stidata$A5MaritalStatus=="5 widowed"|
                                 stidata$A5MaritalStatus=="3 co-habiting"|
                                 stidata$A5MaritalStatus=="4 divorcee"]=0

stidata$Marital_status<-factor(stidata$Marital_status, levels = c(1,0), 
                                       labels = c("Married", "Single"))
table(stidata$Marital_status)

##creating education level
table(stidata$A4LevelOfEducation)
class(stidata$A4LevelOfEducation)
stidata$education_level<-NA                                 
stidata$education_level[stidata$A4LevelOfEducation=="1 none"|
                                stidata$A4LevelOfEducation=="2 primary"]=1;
stidata$education_level[
  stidata$A4LevelOfEducation=="3 secondary"| 
    stidata$A4LevelOfEducation=="4 tertiary"]=0;

stidata$education_level<-factor(stidata$education_level, 
                                      levels = c(1,0),
                                      labels = c("Primary & Below","Secondary & Above"))
table(stidata$education_level)
```

```{r}
#SAVING CLEAN DATASET
write.csv(stidata, "stidata_cleaned.csv", row.names = FALSE)
stidata <- read.csv("stidata_cleaned.csv")
```

# 2. Sociodemographic characteristics of study participants

```{r}
table(stidata$AgeCat)
```

```{r}
library(dplyr)
library(knitr)
```

```{r}
##creating frequency tables
create_table1 <- function(var, var_name) {
  tbl <- table(var)
  pct <- prop.table(tbl) * 100
  data.frame(
    Variable = var_name,
    Category = names(tbl),
    Frequency = as.vector(tbl),
    Percentage = round(as.vector(pct), 1)
  )
}
```

```{r}
# Create tables for each variable
t_age       <- create_table1(stidata$AgeCat, "Age Group")
t_Sex.1       <- create_table1(stidata$Sex.1, "Sex")
t_occupation <- create_table1(stidata$Occupation, "Occupation")
t_church    <- create_table1(stidata$Religion, "Church")
t_education <- create_table1(stidata$education_level, "Education Level")
t_marital   <- create_table1(stidata$Marital_status, "Marital Status")
t_BMI       <- create_table1(stidata$BMI, "BMI")
```

```{r}
# Combine all tables into one master table
table1 <- bind_rows(
  t_age, t_Sex.1, t_occupation, t_church, t_education, t_marital, t_BMI
)
```

```{r}
#Function to add total row for each group
add_totals <- function(df) {
  total_row <- data.frame(
    Variable = unique(df$Variable),
    Category = "Total",
    Frequency = sum(df$Frequency),
    Percentage = round(sum(df$Percentage), 1)
  )
  bind_rows(df, total_row)
}
```

```{r}
str(t_age)
str(t_Sex.1)
str(t_occupation)
str(t_church)
str(t_education)
str(t_marital)
str(t_BMI)
```

```{r}

# Display the table
kable(table1, caption = "Table 1: Sociodemographic Characteristics of Participants")
```

**Interpretation** Most participants were aged 16–34 years (85.8%) with a near-equal split between males (52%) and females (48%). The majority were employed (60%), believers (69.8%), and had secondary education or higher (88%). Most were married (81.3%). In terms of BMI, 43.1% had normal weight, 40.9% were underweight, and a smaller proportion were overweight or obese.

```{r}
#Histogram for Age
library(ggplot2)
ggplot(stidata, aes(x = A1Age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()
```

### Interpretation

The histogram shows the age distribution of participants in the study. The majority of individuals fall between 20 and 35 years, with a peak around the 25–30-year age group. This suggests that most respondents are young adults. The number of participants sharply declines after age 35, with very few individuals above 50 years. This right-skewed distribution indicates a younger population dominates the dataset, which may influence the observed patterns in STI prevalence and associated risk behaviors.

```{r}
table(stidata$Sex.1)
```

```{r}

# Create frequency table from stidata$Sex.1
t_Sex.1 <- as.data.frame(table(Category = stidata$Sex.1))

# Rename columns
colnames(t_Sex.1) <- c("Category", "Frequency")

# Add percentages
t_Sex.1$Percentage <- round(100 * t_Sex.1$Frequency / sum(t_Sex.1$Frequency), 1)

```

```{r}

# Prepare data for pie
Sex.1_pie <- t_Sex.1 %>%
  filter(!is.na(Category)) %>%
  mutate(label = paste0(Category, " (", Percentage, "%)"))

```

```{r}
# Prepare data for pie chart
t_Sex.1 <- stidata %>%
  group_by(Sex.1) %>%
  summarise(Frequency = n()) %>%
  mutate(
    Percentage = round(Frequency / sum(Frequency) * 100, 1),
    Variable = "Sex",
    Category = Sex.1
  ) %>%
  select(Variable, Category, Frequency, Percentage)
```

```{r}
# Create data for pie chart
Sex.1_pie <- t_Sex.1 %>%
  filter(!is.na(Category)) %>%
  mutate(label = paste0(Category, " (", Percentage, "%)"))
```

```{r}
# Plot pie chart
ggplot(Sex.1_pie, aes(x = "", y = Frequency, fill = Category)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  labs(title = "Sex Distribution") +
  theme_void() +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5)) +
  scale_fill_brewer(palette = "Set3")
```

**Interpretation**

The pie chart shows that 52% of respondents are male and 48% are female. This suggests a nearly equal gender distribution, which supports balanced analysis across sexes.

# 3. Prevalence of sexually transmitted infections (STI case status)

```{r}
##prevalence of sexually transmitted infections (STI case status)
class(stidata$CaseStatus)
table(stidata$CaseStatus)
stidata<-stidata[order(-stidata$CaseStatus),]
View(stidata[stidata$CaseStatus==3,])
stidata$CaseStatus[stidata$IdNumber==1 & stidata$A1Age==30]=1
stidata$CaseStatus[stidata$IdNumber==31 & stidata$A1Age==23]=1
table(stidata$CaseStatus)
stidata$STI<-NA
stidata$STI[stidata$CaseStatus==1]=1
stidata$STI[stidata$CaseStatus==2]=0
stidata$STI<-factor(stidata$STI, 
                                 levels = c(1,0),
                                 labels = c("Positive", "Negative"))
table(stidata$STI)

```

```{r}
# Frequency distribution of STI status
prop.table(table(stidata$STI)) * 100
# Pie chart
sti_freq <- table(stidata$STI)

pie(sti_freq,
    col = c("maroon", "blue"),
    main = "Prevalence of Sexually Transmitted Infections",
    labels = paste(names(sti_freq), "\n", round(100 * sti_freq / sum(sti_freq), 1), "%"))
```

**Interpretation**

The prevalence of sexually transmitted infections (STIs) among the study population is approximately 49.8% positive and 50.2% negative. This indicates that STI infection is almost evenly distributed across the sampled individuals. The results highlight the need for robust public health education and prevention programs, as about half of the population is affected

# 4. Association between STI status and other factors

```{r}

names(stidata)
# Load required packages
library(gtsummary)
library(dplyr)
library(gt)  # for nice display in Viewer
```

```{r}
# Create Table 2: Sociodemographic Characteristics and STI Status
stidata %>%
  select(STI, Sex.1, AgeCat, Marital_status, education_level, Occupation, Religion) %>%
  tbl_summary(
    by = STI,            
    missing = "no",      
    percent = "row"     
  ) %>%
  add_overall() %>%     
  add_p() %>%            
  modify_caption("**Table 2: Sociodemographic Characteristics and STI Status**") %>%
  bold_labels() %>%
  as_gt()            
```

**Interpretation:**\

Unemployment was significantly associated with STI status (*p = 0.050*), with a higher STI prevalence among the unemployed (58%). Other variables like sex, age, education level, and religion did not show significant associations (all *p \> 0.05*).

```{r}
# Table 3: Social Capital and STI Status
stidata %>%
  select(STI, Belong, ReceiveHelp, D1BurialSociety, D1religiousgrp, D1savingsClub, 
         D1tradersAssoc, D3FuneralAssistance, D3receivecredit, D3Education) %>%
  tbl_summary(
    by = STI,
    missing = "no",
    percent = "row"
  ) %>%
  add_overall() %>%
  add_p() %>%
  modify_caption("**Table 3: Social Capital and STI Status**") %>%
  bold_labels() %>%
  as_gt()

```

**Interpretation:**\

Receiving help and funeral assistance were significantly associated with lower STI prevalence (*p = 0.028* and *p = 0.019*, respectively). Other forms of social group involvement did not show significant relationships with STI infection.

```{r}
# Table 4: Risky Sexual Behavior and STI Status
stidata %>%
  select(STI, N2SexDebut, N10givereceiveforsex, N11Usedcondom, N12UseCondom, 
         N13TakenAlcohol, N14DoYouHave, SexPartner1year, SexPartner3month, SexPartnerLife3) %>%
  tbl_summary(
    by = STI,
    missing = "no",
    percent = "row"
  ) %>%
  add_overall() %>%
  add_p() %>%
  modify_caption("**Table 4: Risky Sexual Behavior and STI Status**") %>%
  bold_labels() %>%
  as_gt()
```

**Interpretation:**\
Significant associations were observed between STI status and alcohol use, and having multiple sexual partners in the last 3 months, year, or lifetime (*all p \< 0.05*). Condom use and early sexual debut did not show significant associations.

# 5. Risk factors for STI infection using multivariable logistic regression \[Table5\]

statistically significant variables

-   Occupation (p = 0.050)

-   ReceiveHelp (p = 0.028)

-   D3FuneralAssistance (p = 0.019)

-   N13TakenAlcohol (p = 0.016)

-   SexPartner1year (p \< 0.001)

```{r}
# Load dplyr
library(dplyr)

```

```{r}
# Fit multivariable logistic regression model
model <- glm(STI ~ Occupation + ReceiveHelp + D3FuneralAssistance +
             N13TakenAlcohol + SexPartner1year + SexPartner3month +
             SexPartnerLife3,
             data = stidata, family = binomial)
```

```{r}
summary(model)
```

```{r}

# Calculate Odds Ratios (OR),CI, P-values
OR <- exp(coef(model))
CI <- exp(confint(model))
p_values <- summary(model)$coefficients[,4]

# Combine into a table
table5 <- data.frame(
  Variable = names(OR),
  OR = round(OR, 2),
  `95% CI Lower` = round(CI[,1], 2),
  `95% CI Upper` = round(CI[,2], 2),
  `p-value` = round(p_values, 4)
)

# Remove intercept
table5 <- table5[-1, ]

# Print the results
print(table5)
```

## Interpretation: Table 5: Multivariable Logistic Regression of Risk Factors for STI Infection

This multivariable logistic regression model assessed the association between several potential risk factors and the likelihood of being diagnosed with a sexually transmitted infection (STI). The dependent variable was STI status (1 = Positive, 0 = Negative).

-   Occupation (Unemployed): Individuals who were unemployed had significantly lower odds of STI infection compared to those employed (OR = 0.26, 95% CI: 0.12–0.56, *p* = 0.0007). This result is statistically significant and suggests employment may be associated with higher STI risk.

-   Receive Help: Those who reported receiving help had reduced odds of STI infection (OR = 0.48), but this was not statistically significant (*p* = 0.6216), indicating no strong evidence of an association in this model.

-   D3FuneralAssistance: This variable had an OR = 0.14, suggesting lower odds of STI infection among those who received funeral assistance, but the association was not statistically significant (*p* = 0.0981).

-   N13TakenAlcohol (Don't know): This category yielded an extreme OR near zero with a very wide confidence interval, indicating unreliable estimation due to sparse data.

-   SexPartner1year: Participants who had more than one sex partner in the past year had significantly lower odds of STI infection (OR = 0.30, 95% CI: 0.09–0.89, *p* = 0.0329). While this result is counterintuitive, it may reflect confounding factors or misclassification.

-   SexPartner3month & SexPartnerLife3: These variables showed lower odds of STI infection but were not statistically significant (*p* \> 0.05).

Only two factors Occupation (Unemployed) and SexPartner1year were statistically significant predictors of STI status at *p* \< 0.05. Being unemployed and having more than one partner in the past year are associated with reduced odds of STI infection.

# 6. Additional data visualization

```{r}
library(ggplot2)
```

```{r}
#box plot of BMI BY SEX
ggplot(stidata, aes(x = Sex.1, y = BMI, fill = Sex.1)) +
  geom_boxplot() +
  labs(title = "Boxplot of BMI by Sex",
       x = "Sex",
       y = "Body Mass Index (BMI)") +
  theme_minimal()
```

**Interpretation**

The boxplot illustrates the distribution of Body Mass Index (BMI) across different sex groups. Both males and females are represented in all BMI categories including underweight, normal, overweight, obese class I, and obese class II+. This indicates that individuals from both sexes experience a wide range of BMI values which indicates that the presence of both undernutrition and overweight/obesity concerns. The distribution suggests that BMI related health issues are not limited to one sex and interventions should be inclusive of both males and females. Although normal BMI is common in both groups, the presence of higher obesity classes especially among females suggests a need for targeted health strategies.

```{r}
##BMI desity plot
ggplot(stidata, aes(x = BMI, fill = Sex.1)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of BMI",
       x = "Body Mass Index (BMI)",
       y = "Density") +
  theme_minimal()
```

Interpretation:

The peak in the "Normal" category indicates that most individuals have a healthy BMI especially among males. However, there is a noticeable spread into the overweight and obese categories with both males and females showing distributions extending into "Obese Class I" and "Obese Class II+". Females appear to have a slightly higher density in the higher BMI ranges compared to males, suggesting a possible greater prevalence of obesity among females in this sample. The warnings suggest that some categories had very few observations, potentially affecting the smoothness of the curves.

```{r}
#Faceted bar plots comparing STI prevalence by age group and sex.
ggplot(stidata, aes(x =AgeCat, fill = STI)) +
  geom_bar(position = "fill") +
  facet_wrap(~Sex.1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "STI Prevalence by Age Group and Sex",
       x = "Age Group",
       y = "Proportion",
       fill = "STI Status") +
  theme_minimal()
```

**Interpretation**

The bar plot displays the distribution of STI status by age group and sex. Among females, the STI positivity rate is slightly higher in the younger age group (16–34 years) compared to the older group (35–64 years). For males, a notable difference is observed: younger males (16–34 years) show a higher proportion of STI-positive cases compared to older males (35–64 years), who have the lowest positivity rate. This suggests that younger individuals, particularly males aged 16–34, may be at a higher risk of STI infection. These trends highlight the importance of age- and sex-specific interventions in STI prevention programs.
